Forecasting the 2024 US General Election

### 2024-11-04 paper/paper.qmd

---
title: "Forecasting the 2024 US General Election"
subtitle: "Analyzing the influence state on election outcome by using 538 Pollster Data with Bayesian hierarchical model, Predicting Harris Victory[need more details in result]"
author: 
  - Tara Chakkithara
  - Sakura Hu
  - Luka Tosic
thanks: "Code and data are available at https://github.com/LukaTosic09/Forecasting-2024-Election"
date: today
date-format: long
abstract: "This paper forecasts the outcome of the 2024 U.S. general election. Using the poll of polls approach, a generalized linear model was developed with data from FiveThirtyEight [more spcific on method, and what is the date cutoff of the data]  The model's prediction is based on the popular vote in the seven key swing states—Arizona, Nevada, Michigan, Georgia, Wisconsin, Pennsylvania, and North Carolina to estimate the overall winner. Results indicate that Harris is predicted to win, by [how much? need to be more spcific on the result]. This suggests that current polling trends may provide reliable insights into the likely election result, though future shifts in voter sentiment could impact the final outcome."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

### Sample Size
```{r}
#| echo: false
#| warning: false
#| message: false

election_data <- read.csv("../data/raw_data/president_polls.csv")

@@

ggplot(real_data, aes(x = sample_size)) +
  geom_density(fill = "#4B8BBE", alpha = 0.6) + 
  labs(title = "Sample Size",
       x = "Sample Size",
       y = "Density") +
  xlim(0, 10000) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"), 
    axis.title.x = element_text(size = 10, face = "plain"),             
    axis.title.y = element_text(size = 10, face = "plain")               
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

swing_states <- c('Arizona', 'Michigan', 'Georgia', 'Nevada', 
                  'North Carolina', 'Pennsylvania', 'Wisconsin')

map_data <- st_read('../data/raw_data/s_05mr24/s_05mr24.shp', quiet = TRUE)

data <- read_csv('../data/raw_data/president_polls.csv')

data <- data |>
  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) |>
  filter(start_date < biden_drop_date) |>
  filter(pollscore < -1) |>
  select(state, pct, candidate_name) |>
  filter(candidate_name %in% c("Donald Trump", "Joe Biden")) |>  
  filter(state %in% swing_states) |>
  group_by(state, candidate_name) |>  
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")

map_data <- map_data |> 
  rename(state = NAME)

wide_data <- data |>
  pivot_wider(names_from = candidate_name, values_from = avg_pct) |>  
  rename(Trump = `Donald Trump`, Biden = `Joe Biden`)

merged_data <- map_data |>
  left_join(wide_data, by = "state")

merged_data <- merged_data |>
  mutate(color_category = case_when(
    Trump > Biden ~ "Donald Trump",
    Biden > Trump ~ "Joe Biden",
    TRUE ~ "Tie" 
  ))

ggplot(data = merged_data) +
  geom_sf(aes(fill = color_category), color = "white", size = 3, show.legend = FALSE) + 
  scale_fill_manual(values = c("Donald Trump" = "red", "Joe Biden" = "blue", "Tie" = "gray"), 
                    name = "Winner") + 
  labs(
    title = "Voter Sentiment During Biden's Presidential Campaign",
    x = NULL,
    y = NULL
  ) +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +  
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"),
    legend.direction = "horizontal", 
    legend.title = element_text(hjust = 0.5)
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

swing_states <- c('Arizona', 'Michigan', 'Georgia', 'Nevada', 
                  'North Carolina', 'Pennsylvania', 'Wisconsin')

map_data <- st_read('../data/raw_data/s_05mr24/s_05mr24.shp', quiet = TRUE)

data <- read_csv('../data/raw_data/president_polls.csv')

data <- data |>
  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) |>
  filter(start_date > biden_drop_date) |>
  filter(pollscore < -1) |>
  select(state, pct, candidate_name) |>
  filter(candidate_name %in% c("Donald Trump", "Kamala Harris")) |>  
  filter(state %in% swing_states) |>
  group_by(state, candidate_name) |>  
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")

map_data <- map_data |> 
  rename(state = NAME)

wide_data <- data |>
  pivot_wider(names_from = candidate_name, values_from = avg_pct) |>  
  rename(Trump = `Donald Trump`, Harris = `Kamala Harris`)

merged_data <- map_data |>
  left_join(wide_data, by = "state")

merged_data <- merged_data |>
  mutate(color_category = case_when(
    Trump > Harris ~ "Donald Trump",
    Harris > Trump ~ "Kamala Harris",
    TRUE ~ "Tie" 
  ))

ggplot(data = merged_data) +
  geom_sf(aes(fill = color_category), color = "white", size = 3, show.legend = FALSE) + 
  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue", "Tie" = "gray"), 
                    name = "Winner") + 
  labs(
    title = "Voter Sentiment after Biden Drops out of the Presidential Race",
    x = NULL,
    y = NULL
  ) +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +  
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"),
    legend.direction = "horizontal", 
    legend.title = element_text(hjust = 0.5)
  )

@@

```{r}
#| echo: false
#| warning: false
#| message: false

data <- read_csv('../data/raw_data/president_polls.csv')

@@

ggplot(data = wide_data, aes(x = end_date)) +
  geom_line(aes(y = `Kamala Harris`, color = "Kamala Harris"), size = 1) + 
  geom_line(aes(y = `Donald Trump`, color = "Donald Trump"), size = 1) + 
  labs(
    title = "Time Series of Average Polling Percentages",
    x = "Date",
    y = "Average Percentage",
    color = "Candidate"
  ) +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "top"
  )

# Data Overview

@@

# Data Cleaning

@@

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

Talk way more about it.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
library(tidyverse)
library(scales)

@@

ggplot(data, aes(x = x, y = density, color = type)) +
    geom_line(size = 1) + 
    labs(title = "Probability Distribution of Donald Trump's Electoral College Seat Percentage",
         x = "Proportion",
         y = "Density") +
    theme_minimal(base_size = 10) + 
    scale_color_manual(values = c("forestgreen", "purple")) + 
    theme(
        plot.title = element_text(hjust = 0.5, size = 10),
        legend.title = element_blank(), 
        legend.position = "top",
        panel.grid.major = element_line(color = "grey80"),  
        panel.grid.minor = element_blank(),
        axis.ticks = element_line(color = "black"), 
        axis.ticks.length = unit(0.25, "cm"),
        axis.title.x = element_text(margin = margin(t = 10)), 
        axis.title.y = element_text(margin = margin(r = 10))  
    ) +

@@

# Appendix 1 {-}

@@

# Appendix 2 {Idealized Survey}

@@

# 2.1 Sampling

@@

# 2.2 Survey Method

@@

# 2.3 Recruitment

@@

# 2.4 Data Validation

@@

# 2.5 Poll Aggregation and Forecasting

@@

# 2.6 Survey Example

@@

# Additional Data Details

# Model Details {#sec-model-details}

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

params <- data.frame(
  shape1 = c(2, 5.67),
  shape2 = c(3, 9.33),
  label = c("Prior: Beta(2, 3)", "Updated: Beta(5.67, 9.33)")
)

@@

ggplot(density_data, aes(x = x, y = density, color = label)) +
  geom_line(size = 1) +
  labs(
    title = "Distribution of Donald Trump's Electoral College Seat Proportion",
    x = "Probability",
    y = "Density",
    color = "Distribution"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Prior: Beta(2, 3)" = "blue", "Updated: Beta(5.67, 9.33)" = "red")) +
  theme(
    plot.title = element_text(size = 10),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 10)
  )
```

>>>>

---
title: "Forecasting the 2024 US General Election"
subtitle: "Analyzing the influence state on election outcome by using 538 Pollster Data with Bayesian hierarchical model, Predicting Harris Victory[need more details in result]"
author: 
  - Tara Chakkithara
  - Sakura Hu
  - Luka Tosic
thanks: "Code and data are available at https://github.com/LukaTosic09/Forecasting-2024-Election"
date: today
date-format: long
abstract: "This paper forecasts the outcome of the 2024 U.S. general election. Using the poll of polls approach, a generalized linear model was developed with data from FiveThirtyEight [more spcific on method, and what is the date cutoff of the data]  The model's prediction is based on the popular vote in the seven key swing states—Arizona, Nevada, Michigan, Georgia, Wisconsin, Pennsylvania, and North Carolina to estimate the overall winner. Results indicate that Harris is predicted to win, by [how much? need to be more spcific on the result]. This suggests that current polling trends may provide reliable insights into the likely election result, though future shifts in voter sentiment could impact the final outcome."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

### Sample Size
The sample size of a poll is the number of people who participated in the poll. Generally, a larger sample size is seen as a better representation of the population. The most common sample size of polls is around 1250, as seen in @fig-sample-size.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-sample-size
#| fig-cap: "Distribution of Poll Sample Sizes"
election_data <- read.csv("../data/raw_data/president_polls.csv")

@@

ggplot(real_data, aes(x = sample_size)) +
  geom_density(fill = "#4B8BBE", alpha = 0.6) + 
  labs(title = "",
       x = "Sample Size",
       y = "Density") +
  xlim(0, 10000) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"), 
    axis.title.x = element_text(size = 10, face = "plain"),             
    axis.title.y = element_text(size = 10, face = "plain")               
  )
```

### Percentage of Votes
Each poll aims to . By examining how these values fluctuate over time and in response to different candidates, we can better understand their impact on public opinion. Currently, national level polls show Kamala Harris with a 46% approval rating compared to Donald Trump's 43%. This highlights the importance of candidate dynamics in shaping voter sentiment across the electorate.

#### Candidates
```{r}
#| echo: false
#| warning: false
#| message: false
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)

# Define swing states
swing_states <- c('Arizona', 'Michigan', 'Georgia', 'Nevada', 
                  'North Carolina', 'Pennsylvania', 'Wisconsin')

# Load map data
map_data <- st_read('../data/raw_data/s_05mr24/s_05mr24.shp', quiet = TRUE)

# Load polling data
data <- read_csv('../data/raw_data/president_polls.csv')

# Define Biden's withdrawal date
biden_drop_date <- as.Date("2024-07-21")

# Process data before Biden's withdrawal
data1 <- data %>%
  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) %>%
  filter(start_date < biden_drop_date, pollscore < -1) %>%
  filter(candidate_name %in% c("Donald Trump", "Joe Biden"), state %in% swing_states) %>%
  group_by(state, candidate_name) %>%
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")

# Prepare map data
map_data <- map_data %>%
  rename(state = NAME)

# Pivot the data to a wide format
wide_data1 <- data1 %>%
  pivot_wider(names_from = candidate_name, values_from = avg_pct) %>%
  rename(Trump = `Donald Trump`, Biden = `Joe Biden`)

# Merge map and polling data
merged_data1 <- map_data %>%
  left_join(wide_data1, by = "state")

# Categorize color based on polling results
merged_data1 <- merged_data1 %>%
  mutate(color_category = case_when(
    Trump > Biden ~ "Donald Trump",
    Biden > Trump ~ "Joe Biden",
    TRUE ~ "Tie"
  ))

ggplot(data = merged_data1) +
  geom_sf(aes(fill = color_category), color = "white", size = 3, show.legend = FALSE) +
  scale_fill_manual(values = c("Donald Trump" = "red", "Joe Biden" = "blue", "Tie" = "gray"), 
                    name = "Winner") +
  labs(
    title = "Voter Sentiment During Biden's Presidential Campaign",
    x = NULL,
    y = NULL
  ) +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"),
    legend.direction = "horizontal", 
    legend.title = element_text(hjust = 0.5)
  )

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-candidates
#| fig-cap: "Voter Sentiment Before & After Biden's Candidacy Withdrawl"
data2 <- data %>%
  mutate(start_date = as.Date(sub("(..)$", "20\\1", start_date), format = "%m/%d/%Y")) %>%
  filter(start_date > biden_drop_date, pollscore < -1) %>%
  filter(candidate_name %in% c("Donald Trump", "Kamala Harris"), state %in% swing_states) %>%
  group_by(state, candidate_name) %>%
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")

# Pivot the data for post-withdrawal analysis
wide_data2 <- data2 %>%
  pivot_wider(names_from = candidate_name, values_from = avg_pct) %>%
  rename(Trump = `Donald Trump`, Harris = `Kamala Harris`)

# Merge map and polling data for post-withdrawal
merged_data2 <- map_data %>%
  left_join(wide_data2, by = "state")

# Categorize color based on polling results post-withdrawal
merged_data2 <- merged_data2 %>%
  mutate(color_category = case_when(
    Trump > Harris ~ "Donald Trump",
    Harris > Trump ~ "Kamala Harris",
    TRUE ~ "Tie"
  ))

# Create the second plot
ggplot(data = merged_data2) +
  geom_sf(aes(fill = color_category), color = "white", size = 3, show.legend = FALSE) +
  scale_fill_manual(values = c("Donald Trump" = "red", "Kamala Harris" = "blue", "Tie" = "gray"), 
                    name = "Winner") +
  labs(
    title = "Voter Sentiment After Biden Drops Out of the Presidential Race",
    x = NULL,
    y = NULL
  ) +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "plain"),
    legend.direction = "horizontal", 
    legend.title = element_text(hjust = 0.5)
  )
```

#### Poll End Date

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-end-date
#| fig-cap: "Voter Sentiment Across Time For Donald Trump and Kamala Harris" 
data <- read_csv('../data/raw_data/president_polls.csv')

@@

ggplot(data = wide_data, aes(x = end_date)) +
  geom_line(aes(y = `Kamala Harris`, color = "Kamala Harris"), size = 1) + 
  geom_line(aes(y = `Donald Trump`, color = "Donald Trump"), size = 1) + 
  labs(
    title = "",
    x = "",
    y = "PCT",
    color = ""
  ) +
  scale_color_manual(values = c("Kamala Harris" = "blue", "Donald Trump" = "red")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "top"
  )
```

@@

@@

## Data Cleaning

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: model
#| fig-cap: "Probability Distribution of Donald Trump's Electoral College Seat Proportion"
library(tidyverse)
library(scales)

@@

ggplot(data, aes(x = x, y = density, color = type)) +
    geom_line(size = 1) + 
    labs(title = "",
         x = "Proportion",
         y = "Density") +
    theme_minimal(base_size = 10) + 
    scale_color_manual(values = c("forestgreen", "purple")) + 
    theme(
        plot.title = element_text(hjust = 0.5, size = 10),
        legend.title = element_blank(), 
        legend.position = "top",
        panel.grid.major = element_line(color = "grey80"),  
        panel.grid.minor = element_blank(),
        axis.ticks = element_line(color = "black"), 
        axis.ticks.length = unit(0.25, "cm"),
        axis.title.x = element_text(margin = margin(t = 10)), 
        axis.title.y = element_text(margin = margin(r = 10))  
    ) +

@@

# Appendix 1

@@

# Appendix 2 : Idealized Survey

@@

## Sampling

@@

## Survey Method

@@

## Recruitment

@@

## Data Validation

@@

## Poll Aggregation and Forecasting

@@

## Survey Example

@@

# Appendix 3 : Model Details

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: alternate-model
#| fig-cap: "Distribution of Donald Trump's Electoral College Seat Proportion"
params <- data.frame(
  shape1 = c(2, 5.67),
  shape2 = c(3, 9.33),
  label = c("Prior: Beta(2, 3)", "Updated: Beta(5.67, 9.33)")
)

@@

ggplot(density_data, aes(x = x, y = density, color = label)) +
  geom_line(size = 1) +
  labs(
    title = "",
    x = "Probability",
    y = "Density",
    color = "Distribution"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("Prior: Beta(2, 3)" = "blue", "Updated: Beta(5.67, 9.33)" = "red")) +
  theme(
    plot.title = element_text(size = 10),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 10)
  )
```
